Extracting Function Signature from R Program Traces

### 2020-05-12 paper/main.tex
matrix discussion

\lstset{language=R}
\definecolor{LightGray}{rgb}{.92,.92,.92}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}
\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  tabsize=2,
  belowskip=0.5em,
  backgroundcolor=\color{LightGray},
  basicstyle=\small\ttfamily,
  emphstyle=,
  keywordstyle=,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  numbers=left
}
\lstdefinestyle{R}{ %
  language=R,
  morekeywords={assign, delayedAssign},
  deletekeywords={env, equal, c, runif, trace, args, exp, t, all},
  breaklines=true
}
\lstdefinestyle{Rin}{ %
  style=R,
  breaklines=false
}

@@

%
%
\subsubsection{Forgoing Matrices}

Another layer of complexity on vectorized primitives is the presence of
``matrices'', which are \AT{more detail}.  To R, a vector is a
single-dimensional matrix, and R coerces vectors to matrices where
appropriate.  That said, if all vectors are matrices, then any computation
that is valid on a matrix should also be valid on a vector.  Restrictions on
matrix-types arguments amount to restrictions on the dimensions of the
argument, which suggests a dependent type system.  We will touch on this
shortly.  \AT{I can try to come up with some numbers, here. I'm open to a
  discussion on how best to word this.}

>>>>

\lstset{language=R}
\definecolor{LightGray}{rgb}{.92,.92,.92}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}
\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  tabsize=2,
  belowskip=0.5em,
  backgroundcolor=\color{LightGray},
  basicstyle=\small\ttfamily,
  emphstyle=,
  keywordstyle=,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  numbers=left,
  showstringspaces=false
}
\lstdefinestyle{R}{ %
  language=R,
  morekeywords={assign, delayedAssign},
  deletekeywords={env, equal, c, runif, trace, args, exp, t, all},
  breaklines=true
}
\lstdefinestyle{Rin}{ %
  style=R,
  breaklines=false
}

@@

%
%
\subsubsection{Matrices}

On the subject of matrices, they represent another layer of complexity on top of vectorized primitives.
Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and many functions automatically coerce vectors to matrices when appropriate (while not codified in the language semantics, many internal functions perform the coercion).
Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices, particularly those in packages which rely on matrices.
For instance, consider the code in Figure~\ref{fig:matrix-example}.

\begin{lstlisting}
rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
                             na.rm = FALSE, ...) {
  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
  # Validate arguments
  # - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
  # Argument 'x':
  if (!is.matrix(x)) {
    .Defunct(msg = sprintf("Argument 'x' is of class %s, but should be a matrix. 
    The use of a %s is not supported, the correctness of the result is not guaranteed. 
    Please update your code accordingly.", sQuote(class(x)[1]), sQuote(class(x)[1])))
  }
  ...
\end{lstlisting}

\caption{Matrix example from the \code{matrixStats} package.}
\label{fig:matrix-example}
\end{center}
\end{figure}

The \code{rowWeightedMeans} function calculated the weighted means of rows or columns of a matrix \code{x}.
This function exhibits good practice in type checking arguments, rightly producing a message to the user indicating that an explicit matrix should be passed.

To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically.
That said, we don't' intend to subsume all validation code, a topic we discuss next.
